Automatically Proving Equivalence by Type-Safe Reflection

### 2017-03-26 paper-CICM-17/content/1-introduction.tex
CICM paper : intro continued

We revisit an example from previous work~\cite{bradytfp07} 
which shows how proof obligations might appear
when we index a type by natural numbers.  We define the types of
$Bit$ and $Binary$ number in Figure~\ref{binarynums}, 
both indexed over the value they represent, expressed as a natural number.

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Bit : Nat $\rightarrow$ Type where
     b0 : Bit Z
     b1 : Bit (S Z)

data Binary : (width : Nat) $\rightarrow$ (value : Nat) $\rightarrow$ Type where
     zero : Binary Z Z
     (#) : Binary w v $\rightarrow$ Bit bit $\rightarrow$ Binary (S w) (bit + 2 * v)
\end{lstlisting}
\end{center}
\caption{Bit and binary number}
\label{binarynums}
\figrule
\end{figure}

It allows representing a
binary number of width zero with the constructor $zero$, 
and representing larger numbers with $\#$.
We
can extend a binary number $bin$ by adding a digit to its right, and the value
represented by this new binary number is two times the value ($v$) represented
by $bin$ plus the value ($bit$) represented by the added digit.

To write a function performing the addition between two binary numbers,
we start with an auxiliary function (Figure~\ref{addbit}) which adds three bits
(the third one plays the role of a carry flag), and produces the two bits of
the result.

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
addBit : Bit x $\rightarrow$ Bit y $\rightarrow$ Bit c $\rightarrow$ (bX ** (bY ** 
              (Bit bX, Bit bY, c + x + y = bY + 2 * bX)))
addBit b0 b0 b0 = (_ ** (_ ** (b0, b0, Refl)))
addBit b0 b0 b1 = (_ ** (_ ** (b0, b1, Refl)))
addBit b0 b1 b0 = (_ ** (_ ** (b0, b1, Refl)))
addBit b0 b1 b1 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b0 b0 = (_ ** (_ ** (b0, b1, Refl)))
addBit b1 b0 b1 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b1 b0 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b1 b1 = (_ ** (_ ** (b1, b1, Refl)))
\end{lstlisting}
\end{center}
\caption{Addition of three bits}
\label{addbit}
\figrule
\end{figure}

This function produces a dependent pair. The first component is the value $bX$
(expressed as a natural number) represented by the first bit produced (of type
$Bit\ bX$), and the second component is another dependent pair. This
second dependent pair is composed of the value $bY$ represented by the second
bit produced, and of a triple containing the two bits produced and a
proof that these two bits correctly encode the result of the addition.
For example, with the line corresponding to the computation
$1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit $b0$ (encoding $0_2$)
and the bit $b1$ (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

It should be pointed that the proof component of this function is established without any effort, as we always give ``Refl" for the proof. Indeed, for each of the eight patterns, a simple computation reduces the given and expected values to the same value, and ``Refl" can therefore be used. We also note that
there is no need to produce any lemma of correctness about $addBit$ afterwards.
The correct by construction style in which it is written already gives the
property that we want: the two bits produced effectively represent the
addition of the three bits given in input.

We finally define the function adding two binary numbers and a carry flag.
This addition works for two binary numbers
expressed with the same number of bits, and it produces a result with one more
bit. This result represents the value $c + x + y$, where $c$ 
represents the value of the input carry bit, and $x$ and $y$ represent
the values of the two binary numbers given in input. Thus, we'd like to
be able to write:

@@

Unfortunately, the types do not match for either of the two cases. 
The result of the first line $adc\ zero\ zero\ carry$ is
expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] but we
provide a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0))).\]
The problem is even worse for the second case where the expected index is 
\[(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\] while we're trying to provide a term indexed over 
\[plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))\]

For
this situation, Idris provides \emph{provisional definitions} with the 
syntax \texttt{?=}. Using a provisional definition here will make
the definition acceptable, but it also generates two proof obligations, one
for each pattern, both of which say that the provided term can be
converted into a term which has the expected type.  For both of these
patterns, it is possible to prove the equality between the expected and
the provided types. These proofs
consist of a series of rewriting steps, where each step uses a property about the
operation $+$ on natural numbers.  For example, the proof for the second
pattern is shown in Figure \ref{adc_lemma_2} (with some detail elided):

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) 
                 bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{lstlisting}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\label{adc_lemma_2}
\figrule
\end{figure}

Thus, this kind of proof consists of a potentially long sequence of rewriting
steps,
where each step uses one of the available properties. Without some
specific automation, this sequence of rewriting must be done by the programmer.
This is time consuming, and perhaps more importantly, even a small
change in the left
or the right hand side of the equality may invalidate the proof.
This mean that a minor change in our
datatypes, or in the definition of $addBit$ or $adc$ will require us to
replace this proof by another one doing the same job.
This kind
of proof is therefore not reusable, because it performs rewritings for
very specific terms.

If we look at the proof we have written by hand, we realize that we only use
the existence of neutral elements (this is the lemma $plusZeroRightNeutral$),
and the associativity and commutativity of $+$. Thus, we're rewriting a term by
using the properties of a commutative monoid.

We would like to let a generic prover for commutative monoid find a proof on its own.
This should be achievable, because algebraic structures like monoids, groups and
rings (commutative or otherwise) have a very useful property, that
every expression can be normalised to a canonical representation. Thus
these kind of proofs can be automatically produced by computing the
normal forms for each side of the equality, and then comparing them using
syntactical equality. When they are in normal
form, being equal is a matter of being syntactically the same entity.

@@

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing
tactics has been done several times, including for the implementation of a
ring solver for the Coq proof assistant, but without
providing any guarantees, contrary to our type-safe reflection. We will
compare our approach with other implementations in Section~\ref{sect:discuss}.

We make the following contributions:

\begin{itemize}
        \item We present a new type-safe reflection mechanism, where the
        reflected terms are indexed over the concrete inputs, thus providing a
        direct way to pull out the proofs, and providing two guarantees.
        The first guarantee is that the reflection of a term is guaranteed to be a
        faithful representation of the term. The second is that any generated
        proof is guaranteed to be a proof of the required property. The basic
        ideas of the technique are first presented in Section~\ref{sect:ideas} on a
        smaller problem with only natural numbers and addition, and are later
        adapted for the hierarchy of tactics in Section~\ref{sect:general}. 
        Section~\ref{sect:reuse} contains some implementation details.

        \item The normalisation procedures are implemented by following a
        correct by construction approach, instead of implementing a
        normalization procedure, and proving afterwards that this function
        is correct. This approach is much more suitable
        for programming languages like Idris. Again, this approach will be
        presented in Section~\ref{sect:ideas} on the smaller problem, 
        and in Sections~\ref{sect:general} and~\ref{sect:reuse}
        for the complete hierarchy of algebraic structures.

        \item We develop a hierarchy of tactics where each tactic reuses the
        rewriting machinery from the structure immediately below in the hierarchy.
        For example,
        simplifying neutral elements is only implemented at the monoid level,
        and each level inheriting from it will reuse it. Re-usability can be quite tricky
        to obtain when we want to reuse a prover from a structure which is less
        expressive. For example, reusing the monoid prover for building the
        group prover is not trivial, since we lose the possibility to express
        negations ($-x$) and subtractions ($x-y$). Some specific encodings
        will be presented in Section~\ref{sect:reuse} for solving this problem.

\end{itemize}

>>>>

We revisit an example from a previous work of Edwin Brady~\cite{bradytfp07} which shows how proof obligations may arise when a type is indexed by natural numbers. In this development, the goal is to implement a formally verified library of binary numbers. In order to ensure the correctness of the methods that will be developed, we define the types $Bit$ and $Binary$ as indexed over the value they represent (expressed as a natural number), following a correct-by-construction approach.

\begin{lstlisting}
data Bit : Nat $\rightarrow$ Type where
     b0 : Bit Z
     b1 : Bit (S Z)

data Binary : (width : Nat) $\rightarrow$ (value : Nat) $\rightarrow$ Type where
     zero : Binary Z Z
     (#) : Binary w v $\rightarrow$ Bit bit $\rightarrow$ Binary (S w) (bit + 2 * v)
\end{lstlisting}

We want to write a function performing the addition between two binary numbers. In order to do so, we start with an auxiliary function, which adds three bits
(the third one plays the role of a carry flag), and produces the two bits of
the result, where the first one is the most significant bit.

\begin{lstlisting}
addBit : Bit x $\rightarrow$ Bit y $\rightarrow$ Bit c $\rightarrow$ (bX ** (bY ** 
         (Bit bX, Bit bY, c + x + y = bY + 2 * bX)))
addBit b0 b0 b0 = (_ ** (_ ** (b0, b0, Refl)))
addBit b0 b0 b1 = (_ ** (_ ** (b0, b1, Refl)))
addBit b0 b1 b0 = (_ ** (_ ** (b0, b1, Refl)))
addBit b0 b1 b1 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b0 b0 = (_ ** (_ ** (b0, b1, Refl)))
addBit b1 b0 b1 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b1 b0 = (_ ** (_ ** (b1, b0, Refl)))
addBit b1 b1 b1 = (_ ** (_ ** (b1, b1, Refl)))
\end{lstlisting}

This function produces a dependent pair. The first component is the value $bX$ (expressed as a natural number) representing the first bit produced (therefore of type $Bit\ bX$), and the second component is another dependent pair. This second dependent pair is composed of the value $bY$ representing the second bit produced, and of a triple containing the two bits produced and a proof that these two bits correctly encode the result of the addition.
For example, on the fifth line, which corresponds to the computation $1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit $b0$ (encoding $0_2$) and the bit $b1$ (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

We finally define the function $adc$ that adds two binary numbers and a carry flag. This addition works for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result represents the value $c + x + y$, where $c$ represents the value of the input carry bit, and $x$ and $y$ represent the values of the two binary numbers given in input. Therefore, we would like to be able to write :

@@

The first pattern performs the addition between the values zero, zero and a carry. In this simple case, we simply have to produce the one-digit value $zero\ \#\ carry$. The second pattern is when the two binary numbers are respectively $(numx\ \#\ bX)$ and $(numy\ \#\ bY)$ where $numx$ and $numy$ are smaller binary numbers and $bX$ and $bY$ are bits added to the right of $numx$ and $numy$. In this case, we start by adding the least significant bits $bX$ and $bY$ with the $carry$, and this is performed by calling the function $addBit$. It will essentially produce 2 bits, the first one representing the carry of the result (called $carry0$), and the second one the least-significant bit of the result (called $lsb$). Once we have this intermediate result, we can call recursively $adc$ on the smaller terms $numx$ and $numy$ with $carry0$ as the carry that we need to propagate, and finally add $lsb$ to the right of this result.

Unfortunately, this straightforward definition is rejected by Idris' type-checker because the types mismatch for both of the two patterns. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type $Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)$ but we provide a term of type $Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))$.
The problem is even worse for the second case where the expected index is \\
$(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))$
while we're trying to provide a term indexed over \\
$plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))$.
We know that the definition of $adc$ we have given is correct, and it has \emph{provably} the expected type, but it does not have it \emph{immediately} or \emph{by reduction} : after full reductions\footnote{In dependently typed theories, the presence of types predicated on values creates the need of using reductions also inside of types}, the expected and provided types are still different, and idris can not figure out a proof of compatibility on its own. 
If we use provisional definitions[ref], the previous definition can be accepted by the system, but we now have two proof obligations $adc\_lemma\_1$ and $adc\_lemma\_2$ demanding proofs of equality between the two types, and they have to be provided by hand --here with a proof script-- :

\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) 
                 bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
}
\end{lstlisting}

As we see, this kind of proof consists of a potentially long sequence of rewriting steps, where each step uses one of the available properties: neutral element, commutativity, associativity. Without some specific automation, this sequence of rewriting must be done by the programmer.
This is time consuming and more importantly, just a small change in the left or the right hand side of the equality may invalidate the proof. That means that a minor change in our datatypes, or in the definition of $addBit$ or $adc$ will require us to replace this proof by another one fulfilling the same task.
This kind of proof is therefore absolutely not reusable, because it performs rewritings for very specific terms.
These proofs become the everyday routine in any dependently-typed language, as soon as we index types over values in order to capture some logical properties.  As we've seen, when a type $T$ is indexed over $Nat$, the definition of a function which has codomain $T$ is likely to require a proof of equality between the produced index and the expected one.

The proof that we have written by hand only uses the existence of neutral elements, and the associativity and commutativity of $+$ on $Nat$. Thus, we're rewriting a term by using the properties of a commutative monoid. We would like to let a generic prover for commutative monoid find a proof on its own.

@@

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing tactics has been done several times, including for the implementation of a ring solver for the Coq proof assistant, but without
providing any guarantees, contrary to our type-safe reflection. We will
compare our approach with other implementations in Section~\ref{sect:AlternativeApproaches}.

The contributions of this paper are the following :

\begin{enumerate}

\item We present a new type-safe reflection mechanism, where the reflected terms are indexed over the concrete inputs, thus providing a
direct way to pull out the proofs, and providing the guarantee that the reflection of a term is indeed a faithful representation of the term.
The basic ideas of the technique are first presented in Section~\ref{sect:aSimpleProofAutomation} on a smaller problem with only natural numbers and addition, and are later adapted for a hierarchy of tactics proving equalities in algebraic strcures, in Section~\ref{sect:ProvingEquivalencesInAlgebraicStructures}.

\item The normalisation procedures are implemented by following a
correct by construction approach, instead of implementing a
normalisation procedure, and proving afterwards that this function
is correct with external lemmas. This approach is much more suitable
for programming languages like Idris. Again, this approach will be
presented in Section~\ref{sect:aSimpleProofAutomation} on the smaller problem, 
and in Sections~\ref{sect:ProvingEquivalencesInAlgebraicStructures} for the complete hierarchy of algebraic structures.

\item We develop a hierarchy of tactics where each tactic reuses the
rewriting machinery of the structure from which it naturally inherits.
For example, simplifying neutral elements is only implemented at the monoid level,
and each level inheriting from it will reuse it. Re-usability is difficult to obtain when we want to reuse the prover of a less expressive structure. For example, reusing the monoid prover for building the
group prover is not trivial, since we lose the possibility to express
negations ($-x$) and subtractions ($x-y$). Some specific encodings
will be presented in Section~\ref{sect:NormalisationsFunctionsAndReusabilityOfTheProvers} for solving this problem.

\end{enumerate}

### 2017-03-26 paper-CICM-17/paper.tex
CICM paper : intro continued

We will first present the technique on a small example, where we aim to automatically prove equalities with universally quantified natural numbers and the associative addition operation, and will later generalise these ideas for building automatic provers for various kind of properties (commutativity, distributivity, existence of neutral elements...), leading to a hierarchy of reflexive tactics for proving equivalences in Semi-Groups, Monoids, Commutative Monoids, Groups, Commutative Groups, Semi-Rings and Rings. We will also show how each tactic reuses the ones from the simplest structures, thus avoiding as much as possible the duplication of code and proofs.
\keywords{proof automation, equivalence, equality, proof by reflection, correct-by-construction software}
\end{abstract}
%

>>>>

After a presentation of our key ideas on a small example, we will build a hierarchy of reflexive tactics for proving equivalences in Semi-Groups, Monoids, Commutative Monoids, Groups, Commutative Groups, Semi-Rings and Rings. We will also show how each tactic reuses the ones from the simplest structures, thus avoiding as much as possible the duplication of code and proofs.
\keywords{proof automation, equivalence, equality, proof by reflection, correct-by-construction software}
\end{abstract}
%
